---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
```{r}
install.packages("matlib")
install.packages("rsample")
```

# Importing needed library
```{r}
library(matlib)
library(ggplot2)
library(rsample)
```

# Importing input data from a csv file
```{r}
X=as.matrix(read.csv(file="C:/Users/Prabin's PC/OneDrive/MSc. Assignment/x.csv",header = F))
head(X)
colnames(X)<-c("x1","x3","x4","x5")
```

# Displaying imported data
```{r}
head(X)
```
# Importing targeted data
```{r}
Y=as.matrix(read.csv(file="C:/Users/Prabin's PC/OneDrive/MSc. Assignment/y.csv",header = F))
colnames(Y)<-c("y")
```

# Displaying targeted data
```{r}
head(Y)
```

# Importing time series data data 
```{r}
time = read.csv("C:/Users/Prabin's PC/OneDrive/MSc. Assignment/t.csv", header = F)
time = as.matrix(time)
```

# Displaying time series data
```{r}
head(time)
```

# Task 1.1
# Plotting time series data

```{r}
X.ts<-ts(X,start = c(min(time),max(time)),frequency =1)
Y.ts<-ts(Y,start = c(min(time),max(time)),frequency =1)
```

# Plotting timeseries data of input and target variable
```{r}
plot(X.ts[400:500,1],main = "Time series plot of X1 Signal", xlab = "Time", ylab = "Input signal", col="#1B8DE4",type="l")
plot(X.ts[400:500,2],main = "Time series plot of X3 Signal", xlab = "Time", ylab = "Input signal", col="#1B8DE4",type="l")
plot(X.ts[400:500,3],main = "Time series plot of X4 Signal", xlab = "Time", ylab = "Input signal", col="#1B8DE4",type="l")
plot(X.ts[400:500,4],main = "Time series plot of X5 Signal", xlab = "Time", ylab = "Input signal", col="#1B8DE4",type="l")

plot(Y.ts[400:500],main = "Time series plot of Y Signal", xlab = "Time", ylab = "Output signal", col="#E4721B" ,type="l")
```
# Task 1.2 : Plotting distribution of each input Power plant energy signal
# Creating a density of all input signal X 

```{r}
density_of_X=density(X)
plot(density_of_X,main = "Density plot of input signal X",col="#1B8DE4")
```

# Creating a Histogram of X signal

```{r}
hist(X,freq = FALSE,main = "Density",col="#e4721b")
```


# Combining Histogram of X signal with density plot
```{r}
hist(X,freq = FALSE,main = "Density",col="#e4721b")
lines(density_of_X,lwd=2,col="#1B8DE4")
rug(jitter(X))
```
# Histogram and density plot of individual input signal X and output signal y

```{r}
# Creating a density plot of input signal X1 
X1_density=density(X[,"x1"])
hist(X[,"x1"],freq = FALSE,main = "Histogram and density plot of Temperature (°C)",xlab = "x1 Signal", col="#e4721b")
lines(X1_density,lwd=2,col="#1B8DE4")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x1"]))

# Creating a density plot of input signal X3 
density_of_X3=density(X[,"x3"])
hist(X[,"x3"],freq = FALSE,main = "Histogram and density plot of Ambient Pressure (millibar)",xlab = "x3 Signal", col="#e4721b")
lines(density_of_X3,lwd=2,col="#1B8DE4")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x3"]))

# Creating a density plot of input signal X4
density_of_X4=density(X[,"x4"])
hist(X[,"x4"],freq = FALSE,main = "Histogram and density plot of Relative Humidity (%)",xlab = "x4 Signal", col="#e4721b")
lines(density_of_X4,lwd=2,col="#1B8DE4")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x4"]))

# Creating a density plot of input signal X5
density_of_X5=density(X[,"x5"])
hist(X[,"x5"],freq = FALSE,main = "Histogram and density plot of Exhaust Vacuum (cm, Hg)",xlab = "x5 Signal", col="#e4721b")
lines(density_of_X5,lwd=2,col="#1B8DE4")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x5"]))

# Creating a density plot of output signal y
density_of_y=density(Y[,"y"])
hist(Y[,"y"],freq = FALSE,main = "Histogram and density plot of y",xlab = "y Signal", col="#e4721b")
lines(density_of_y,lwd=2,col="#1B8DE4")
# Add the data-points with noise in the X-axis
rug(jitter(Y[,"y"]))


```

# Task 1.3 Creating scatter plots to indeitify correlation
# Arranging plot in a single screen

```{r}
par(mfrow=c(2,2))

# Plotting input signal Temperature against output signal Y
plot(X[,"x1"],Y,main = "Correlation betweeen Temperature and Y signal", xlab = "X1 signal", ylab = "Output signal y", col="#1B8DE4")

# Plotting input signal Ambient Pressure against output signal Y
plot(X[,"x3"],Y,main = "Correlation betweeen Ambient Pressure and Y signal", xlab = "X3 signal", ylab = "Output signal y", col="#e4721b")

# Plotting input signal Relative Humidity against output signal Y
plot(X[,"x4"],Y,main = "Correlation betweeen Relative Humidity and Y signal", xlab = "X4 signal", ylab = "Output signal y", col="#1B8DE4")

# Plotting input signal X5 against output signal Y
plot(X[,"x5"],Y,main = "Correlation betweeen Exhaust Vaccuum and Y signal", xlab = "X5 signal", ylab = "Output signal y", col="#e4721b")

```
# Task 2
# Calculating ones for binding the data
```{r}
ones = matrix(1 , length(X)/4,1)
head(ones)
```
# Task 2.1
# Calculating thetahat of each candidate model
```{r}
# Binding data from equation of model 1.
X_model1<-cbind(ones,X[,"x4"], X[,"x3"]^2)
head(X_model1)
# Calculating thetahat of Model 1
Model1_thetahat=solve(t(X_model1) %*% X_model1) %*% t(X_model1) %*% Y
Model1_thetahat

# For model 2
#Binding data from equation of model 2.
X_model2<-cbind(ones,X[,"x4"],X[,"x3"]^2,X[,"x5"])
head(X_model2)
# Calculating thetahat of Model 2
Model2_thetahat=solve(t(X_model2) %*% X_model2) %*% t(X_model2) %*% Y
Model2_thetahat

# For Model 3
#Binding data from equation of model 3.
X_model3<-cbind(X[,"x3"],X[,"x4"],X[,"x5"]^3)
head(X_model3)
# Calculating thetahat of Model 3
Model3_thetahat=solve(t(X_model3) %*% X_model3) %*% t(X_model3) %*% Y
Model3_thetahat

# For model 4
#Binding data from equation of model 4.
X_model4<-cbind(ones,X[,"x4"],(X[,"x3"])^2,scale((X[,"x5"])^3))
head(X_model4)
# Calculating thetahat of Model 4
Model4_thetahat=qr.solve(t(X_model4) %*% X_model4) %*% t(X_model4) %*% Y
Model4_thetahat

# For Model 5
# Binding data from equation of model 5.
X_model5<-cbind(ones,(X[,"x4"]),(X[,"x1"])^2,(X[,"x3"])^2)
head(X_model5)
# Calculating thetahat of model 1
Model5_thetahat=solve(t(X_model5) %*% X_model5) %*% t(X_model5) %*% Y
Model5_thetahat

```
# Printing value of theta of each model
```{r}
#model1
Model1_thetahat
t(Model1_thetahat)
#model 2
Model2_thetahat
t(Model2_thetahat)
#model 3
Model3_thetahat
t(Model3_thetahat)
#model 4
Model4_thetahat
t(Model4_thetahat)
#model 5
Model5_thetahat
t(Model5_thetahat)
```
# Task 2.2
# Calculating Y-hat and RSS for each model
```{r}
# Calculating Y-hat and RSS Model 1
Y_hat_model1 = X_model1 %*% Model1_thetahat
head(Y_hat_model1)
# Calculating RSS
RSS_Model_1=sum((Y-Y_hat_model1)^2)
head(RSS_Model_1)

# Calculating Y-hat and RSS of model 2
Y_hat_model2 = X_model2 %*% Model2_thetahat
head(Y_hat_model2)
# Calculating RSS
RSS_Model_2=sum((Y-Y_hat_model2)^2)
head(RSS_Model_2)

# Calculating Y-hat and RSS of model 3
Y_hat_model3 = X_model3 %*% Model3_thetahat
head(Y_hat_model3)
# Calculating RSS
RSS_Model_3=sum((Y-Y_hat_model3)^2)
head(RSS_Model_3)
 
# Calculating Y-hat and RSS of model 4
Y_hat_model4 = X_model4 %*% Model4_thetahat
head(Y_hat_model4)
# Calculating RSS
RSS_Model_4=sum((Y-Y_hat_model4)^2)
head(RSS_Model_4)

# Calculating Y-hat and RSS of model 5
Y_hat_model5 = X_model5 %*% Model5_thetahat
head(Y_hat_model5)
# Calculating RSS
RSS_Model_5=sum((Y-Y_hat_model5)^2)
head(RSS_Model_5)
```

# Printing RSS value

```{r}
model1 <- c(RSS_Model_1)
model2 <- c(RSS_Model_2)
model3 <- c(RSS_Model_3)
model4 <- c(RSS_Model_4)
model5 <- c(RSS_Model_5)

dfRSS <- data.frame(model1, model2,model3,model4,model5)
dfRSS

#anova(lm(X_model2), lm(X_model5))
#anova
```

# Task 2.3 Calculating likelihood and Variance of each model
```{r}
N=length(Y)

# Calculating the Variance of Model 1
Variance_model1=RSS_Model_1/(N-1)
Variance_model1

# Calculating the log-likelihood of Model 1
likehood_Model_1=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model1))-(1/(2*Variance_model1))*RSS_Model_1
likehood_Model_1

# Calculating Variance and log-likelihood of Model 2
Variance_model2=RSS_Model_2/(N-1)
Variance_model2
likehood_Model_2=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model2))-(1/(2*Variance_model2))*RSS_Model_2
likehood_Model_2


# Calculating Variance and log-likelihood of Model 3
Variance_model3=RSS_Model_3/(N-1)
Variance_model3
likehood_Model_3=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model3))-(1/(2*Variance_model3))*RSS_Model_3
likehood_Model_3

# Calculating Variance and log-likelihood of Model 4
Variance_model4=RSS_Model_4/(N-1)
Variance_model4
likehood_Model_4=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model4))-(1/(2*Variance_model4))*RSS_Model_4
likehood_Model_4

# Calculating Variance and log-likelihood of Model 5
Variance_model5=RSS_Model_5/(N-1)
Variance_model5
likehood_Model_5=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model5))-(1/(2*Variance_model5))*RSS_Model_5
likehood_Model_5
```
# Printing variance  values

```{r}
model1 <- c(Variance_model1)
model2 <- c(Variance_model2)
model3 <- c(Variance_model3)
model4 <- c(Variance_model4)
model5 <- c(Variance_model5)

dfVariance <- data.frame(model1, model2,model3,model4,model5)
dfVariance
```

# Printing likelihood values

```{r}
model1 <- c(likehood_Model_1)
model2 <- c(likehood_Model_2)
model3 <- c(likehood_Model_3)
model4 <- c(likehood_Model_4)
model5 <- c(likehood_Model_5)

dfLikelihood <- data.frame(model1, model2,model3,model4,model5)
dfLikelihood
```

# Task 2.4 
# Calculating AIC And BIC of each model
```{r}
# Calculating AIC and BIC of model 1
K_model1<-length(Model1_thetahat)
K_model1
AIC_model1=2*K_model1-2*likehood_Model_1
AIC_model1
BIC_model1=K_model1*log(N)-2*likehood_Model_1
BIC_model1

## thetahat of model 2
K_model2<-length(Model2_thetahat)
K_model2
##Calculating AIC and BIC of model 2
AIC_model2=2*K_model2-2*likehood_Model_2
AIC_model2
BIC_model2=K_model2*log(N)-2*likehood_Model_2
BIC_model2

## thetahat of model 3
K_model3<-length(Model3_thetahat)
K_model3
##Calculating AIC and BIC of model 3
AIC_model3=2*K_model3-2*likehood_Model_3
AIC_model3
BIC_model3=K_model3*log(N)-2*likehood_Model_3
BIC_model3

## thetahat of model 4
K_model4<-length(Model4_thetahat)
K_model4
##Calculating AIC and BIC of model 4
AIC_model4=2*K_model4-2*likehood_Model_4
AIC_model4
BIC_model4=K_model4*log(N)-2*likehood_Model_4
BIC_model4

## thetahat of model 5
K_model5<-length(Model5_thetahat)
K_model5
##Calculating AIC and BIC of model 5
AIC_model5=2*K_model5-2*likehood_Model_5
AIC_model5
BIC_model5=K_model5*log(N)-2*likehood_Model_5
BIC_model5
```
# Printing K values

```{r}
model1 <- c(K_model1)
model2 <- c(K_model2)
model3 <- c(K_model3)
model4 <- c(K_model4)
model5 <- c(K_model5)

dfK <- data.frame(model1, model2,model3,model4,model5)
dfK
```
# Printing AIC values

```{r}
model1 <- c(AIC_model1)
model2 <- c(AIC_model2)
model3 <- c(AIC_model3)
model4 <- c(AIC_model4)
model5 <- c(AIC_model5)

dfAIC <- data.frame(model1, model2,model3,model4,model5)
dfAIC
```

# Printing BIC values

```{r}
model1 <- c(BIC_model1)
model2 <- c(BIC_model2)
model3 <- c(BIC_model3)
model4 <- c(BIC_model4)
model5 <- c(BIC_model5)

dfBIC <- data.frame(model1, model2,model3,model4,model5)
dfBIC
```

## Task 2.5 calculating error plotting normal/gaussian distibution of each plot

```{r}
par(mfrow=c(1,1))

## Error of model1
model1_error <- Y-Y_hat_model1
head(model1_error)

## Plotting the graph QQplot and QQ line of model 1
qqnorm(model1_error, col = "#e4721b",main = "QQ plot of model 1")
qqline(model1_error, col = "#1B8DE4",lwd=1)

## Error of model2
model2_error <- Y-Y_hat_model2 # error of model 2
## Plotting QQplot and QQ line of model 2
qqnorm(model2_error, col = "#e4721b",main = "QQ plot of model 2")
qqline(model2_error, col = "#1B8DE4")

## Error of model3
model3_error <- Y- Y_hat_model3
## Plotting QQplot and QQ line of model 3
qqnorm(model3_error, col = "#e4721b",main = "QQ plot of model 3")
qqline(model3_error, col = "#1B8DE4")

## Error of model4
model4_error <- Y-Y_hat_model4
## Plotting QQplot and QQ line of model 4
qqnorm(model4_error, col = "#e4721b",main = "QQ plot of model 4")
qqline(model4_error, col = "#1B8DE4")

## Error of model5
model5_error <- Y- Y_hat_model5
## Plotting QQplot and QQ line of model 5
qqnorm(model5_error, col = "#e4721b",main = "QQ plot of model 5")
qqline(model5_error, col = "#1B8DE4")
```

# Task 2.7 splitting data into training and testing dataset and calculating estamation based on training dataset
#also plotting normal distribution graph of training data
```{r}
## Spliting the dataset y into  Traning and testing data set.
split_Y<-initial_split(data = as.data.frame(Y),prop=.7)
## Traning splitted Y dataset 
Y_training_set<-training(split_Y)
Y_testing_set<-as.matrix(testing(split_Y))
## Testing splitted Y dataset 
Y_training_data<-as.matrix(Y_training_set)

## Spliting the dataset of X into  Traning and testing data set.
split_X<-initial_split(data = as.data.frame(X),prop=.7)
## Traning splitted X dataset
X_training_set<-training(split_X)
## Testing splitted X dataset 
X_testing_set<-as.matrix(testing(split_X))
X_testing_data<-as.matrix(X_testing_set)
X_training_data<-as.matrix(X_training_set)

### Estimating model parameters using Traning set
traning_ones=matrix(1 , length(X_training_set$x1),1)
# selected model 5 and using equation of model 5
X_traning_model<-cbind(traning_ones,X_training_set[,"x4"],(X_training_set[,"x1"])^2,(X_training_set[,"x3"])^2)
traning_thetahat=Ginv(t(X_traning_model) %*% X_traning_model) %*% t(X_traning_model) %*%  Y_training_data
  
### Model out/Prediction
Y_testing_hat = X_testing_data %*% traning_thetahat
head(Y_testing_hat)
RSS_testing=sum((Y_testing_set-Y_testing_hat)^2)
head(RSS_testing)
t.test(Y_training_data, mu=500, alternative="two.sided", conf.level=0.95)
C_I1=454.0137
C_I2=454.8377
p2 <- plot(density(Y_training_data), col="#1B8DE4", lwd=2,
           main="Distribution of Traning Data")
abline(v=C_I1,col="#e4721b", lty=2)
abline(v=C_I2,col="#e4721b", lty=2)

thetaHat_training =solve(t(X_training_data) %*% X_training_data) %*% t(X_training_data) %*%Y_training_data
head(thetaHat_training)
length(thetaHat_training)
dis_test=density(Y_training_data)
plot(dis_test,main = "Density plot of Y Signal",col="#1B8DE4")

### Calculating Confidential interval
 z=1.96 ##(95%) Confidential interval
 error=((Y_testing_set-Y_testing_hat))
 n_len=length(Y_testing_hat)
 C_I_1= z * sqrt( (error * (1-error) ) / n_len)
 head(C_I_1)
 head(error)
 C_I_2= z * sqrt( (error * (1+error)) / n_len)
 head(C_I_2)        

Variance_model = mean((Y_testing_set - Y_testing_hat)^2)
se = sqrt(Variance_model)
CI_lower = Y_testing_hat - z * se
CI_upper = Y_testing_hat + z * se
head(CI_lower)
head(CI_upper)
```

# Task 3
```{r}
## Model 5 will be used, parameter are selected and kept constant.
arr_1=0
arr_2=0
f_value=0
s_value=0
Model5_thetahat
#values from thetahat
thetebias <- 251.671653135 #selected parameter
thetaone <- 0.227168908 # selected parameter
thetatwo <- -0.036521863 # constant value
thetathree <- -0.003650664 # constant value

 Epison <- RSS_Model_5 * 2 ## fixing value of eplision
num <- 100 #number of iteration
##Calculating Y-hat for performing rejection ABC
counter <- 0
for (i in 1:num) {
  range1 <- runif(1, thetebias * 0.8, thetebias * 1.2)
  range2 <- runif(1, thetaone * 0.8, thetaone * 1.2)
  New_thetahat <- matrix(c(range1,range2,thetatwo,thetathree))
  New_Y_Hat <- X_model5 %*% New_thetahat ## calculating new Y-hat
  new_RSS <- sum((Y-New_Y_Hat)^2)
  new_RSS
  if (new_RSS > Epison){
    arr_1[i] <- range1
    arr_2[i] <- range2
    counter = counter+1
    f_value <- matrix(arr_1)
    s_value <- matrix(arr_2)
  }
}
hist(f_value, col = "#e4721b")
hist(s_value, col = "#1B8DE4")


```
```{r}
# 1. Acceptance Rate
acceptance_rate <- counter / num
cat("Acceptance Rate:", acceptance_rate, "\n")

# 2. Accepted Parameters
# Filter non-zero values (accepted samples)
accepted <- arr_1 != 0 & arr_2 != 0
accepted_params <- data.frame(range1 = arr_1[accepted], range2 = arr_2[accepted])
cat("Number of Accepted Samples:", nrow(accepted_params), "\n")
if (nrow(accepted_params) > 0) {
  print("Accepted Parameters (range1, range2):")
  print(accepted_params)
} else {
  cat("No samples were accepted.\n")
}

# 3. Summary Statistics
if (nrow(accepted_params) > 0) {
  cat("\nSummary Statistics for range1:\n")
  print(summary(accepted_params$range1))
  cat("\nSummary Statistics for range2:\n")
  print(summary(accepted_params$range2))
}

# 4. Visualizations
# Scatter plot of accepted range1 vs. range2
if (nrow(accepted_params) > 0) {
  plot(accepted_params$range1, accepted_params$range2,
       xlab = "range1 (Intercept)", ylab = "range2 (x4_scaled)",
       main = "Accepted Parameters", pch = 19, col = "#e4721b")
  abline(v = thetebias, h = thetaone, col = "#1B8DE4", lty = 2)  # Original values
}

# Histograms
if (nrow(accepted_params) > 0) {
  par(mfrow = c(1, 2))
  hist(accepted_params$range1, main = "Histogram of range1", xlab = "range1", col = "#e4721b")
  abline(v = thetebias, col = "#1B8DE4", lty = 2)
  hist(accepted_params$range2, main = "Histogram of range2", xlab = "range2", col = "#e4721b")
  abline(v = thetaone, col = "#1B8DE4", lty = 2)
  par(mfrow = c(1, 1))
}
```


```{r}
### Ploting Joint and Marginal Posterior Distribution of the graph
plot(f_value,s_value, col = c("#1B8DE4", "#e4721b"), main = "Joint and Marginal Posterior Distribution")
par(mfrow=c(1,1))
```


